# Chapter 01

## Feature Engineering and Selection (by Kuhn and Johnson)

Main reference: see <http://www.feat.engineering/>

```{r}
# ------------------------------------------------------------------------------
# Feature Engineering and Selection: A Practical Approach for Predictive Models
# by Max Kuhn and Kjell Johnson
# Code for Section 1.1 at
# https://bookdown.org/max/FES/intro-intro.html#a-simple-example
#
# ------------------------------------------------------------------------------
# 
# Code requires these packages: 
library(caret)
library(tidymodels)
library(ggthemes)
```

## EDA

Create sample data from Chapter 01:

```{r}
# Create example data ----------------------------------------------------------
data(segmentationData)
```

View/inspect the dataframe to get a sense of how it looks:

```{r}
#View(segmentationData)
```

Note: one could use the `skimr` package to compute summary statistics as well: 

```{r}

```



**Split the dataset**
This dataframe comes with a label predefined indicating if a sample is used for testing or training.


```{r}
# rename some predictors
# segmentationData

```

Get training/testing data: 

```{r}
# define training data


# define testing data

```



**Prepare to train a model**

Simple example of logistic regression using natural units (this uses functions from the `caret` package to train models)

```{r}
# Do analysis in natural units -------------------------------------------------

```


A first plot of: 

```{r}

```


## Modeling


Train model using natural terms: 


```{r}

```

Run some predictions on testing data

```{r}

```



Create the ROC plot

```{r}

```

## Refine the Model

Analysis on transformed data:

```{r}
# Analysis of transformed data -------------------------------------------------

```


Get predictions in testing dataset using refined model:

```{r}

```

Generate plot that uses transformed predictors:

```{r}

```

## Model Comparison 

Combined ROC plots: 


```{r}
# Combine dataframe for plot -------------------------------------------------------------

```

Reproduce plot from FES book

```{r}
# https://bookdown.org/max/FES/a-simple-example.html#fig:intro-natural

```

