---
title: "Clustering Lab"
output: html_notebook
---


# Libraries

```{r}
library(ISLR2)
library(factoextra)
library(ggplot2)
```


```{r}
data(iris)
head(iris)
```


# K-means Clustering

Since we are wanting to work with numerical data, we want to drop the Species column
```{r}
iris_Num <- iris[,-5]
```

Next, we can use the `kmeans()` function to run k-means clusters with k = 2 groups
```{r}
K_Iris <- kmeans(iris_Num, centers = 2)
```

```{r}
K_Iris$cluster
```

NOTE: we can run these over and over and we will get different results every time because it is random! 

```{r}
fviz_cluster(K_Iris, data = iris_Num,
             palette = c("pink", "#E7B800"), 
             geom = "point",
             ellipse.type = "convex", 
             ggtheme = theme_bw()
             ) 
```

NOW, we can go back and use different Ks to see what we got. 
NOTE: the big Circle and big Triangle are the centroids

```{r}
K_Iris$centers
```

Neat trick:

Sometimes I want to know more about how these were clustered, so I might combine the data sets to make a new data set

```{r}
MIX <- bind_cols(iris, K_Iris$cluster)
```

```{r}
table(MIX$...6, MIX$Species)
```

Here is a neat little table that gives you an idea of how the clustered ended up by species, which might be a good indicator of how well your clusters worked.

# Hierarchical Clustering

```{r}
#Type dist to see what the purpose is in this clustering method

```


```{r}
dis <- dist(iris_Num)
```

Now complete hierarchical clustering: 

```{r}
Complete <- hclust(dis, method = 'complete')
plot(Complete, main = "Complete Linkage", cex = 0.1)
```


NOW lets standardize the data and see how that might have changed things:
```{r}
Standard_iris_Num <-  scale(iris_Num, center = T, scale = T) #standardizing
Standard_Complete <- hclust(dist(Standard_iris_Num), method = "complete") #new cluster
plot(Standard_Complete, main = "Standardized Complete Linkage", cex = 0.1) #plotting
rect.hclust(Standard_Complete, k=3, border=2:5) #making clusters of k=3 visually
```

Of course we can do all this with different linkages:

Single:

```{r}
Standard_Single <- hclust(dist(Standard_iris_Num), method = "single") #new cluster
plot(Standard_Single, main = "Standardized Single Linkage", cex = 0.1) #plotting
```

Average:

```{r}
Standard_Average <- hclust(dist(Standard_iris_Num), method = "average") #new cluster
plot(Standard_Average, main = "Standardized Average Linkage", cex = 0.1) #plotting
```









