---
title: "Homework 4"
output: html_notebook
author: "Halan Badilla Osorio"
---

What I want you to do for this assignment is pick just one of these models and do a little light research. 

1. describe how this model works

The simplest way I can think of explaining this is that this is you using regression but you are splitting your x-var into intervals allowing you to do multiple regressions that best fit in their own separate intervals.

Good link that goes really in depth with them. It is for Python though.
https://diogoribeiro7.github.io/data%20science/understanding_splines_what_they_how_they_used_data_analysis/

2. what type of data is this model best used for

Technically this as far as I understand can be used in anything in which you have some sort of regression you're applying. Specifically though, you'd want to use Splines when you are working with *non-linear trends*, *if your data has any sort of threshold/milestone*, *periodic data* and if *you are missing data and want to interpolate or fill a gap.* 

3. what are the advantages and disadvantages

Pros:
-Can graph non-linear data better
-Flexible
-Interpretability (Clear segmentation)

Cons:
-You need lots of data, can cause you to overfit
-Complex
-You need to place knots
-Interpretability (Can be a con if it gets too complicated and with the more complicated splines)

4. make a model (it doesn't have to be complex, just an example)

```{r}
Sys.setenv(LANG = "en")
library(dplyr)
library(tidyverse)
library(lubridate)
library(ggplot2)
```

```{r}
incidents <- incidents %>% # This will add 3 columns separating the parts of the date
  mutate(
    year = year(date),
    month = month(date, label = TRUE),
    day = day(date)
  )

incidents <- incidents %>% # This will make a separate column of year and month
  mutate(
    yearMonth = make_date(year, month) %>%
      format("%Y-%m")
  )

incidents <- incidents %>% # Sum the total of deaths per month
  group_by(yearMonth) %>%
  mutate(deathsByMonth = sum(num_victims_killed))

incidents <- incidents %>% # Sum the total of injuries per month
  group_by(yearMonth) %>%
  mutate(injuredByMonth = sum(num_victims_injured))

victimIncidentMerge <- merge(incidents, victims, key = incident_id)
```

```{r}
simpleLinearDeaths <- lm(age ~ as.factor(race) + as.factor(sex), victims)
summary(simpleLinearDeaths)
```

```{r}
ggplot()
```

5. explain the output and how well the model functioned
6. cite a research paper which uses this model type in its work